Vagueness, Noise, and Signaling

### 2015-09-29 paper/version_04/paper.tex
Small fixes (cherry picked from commit 7d7fc88)

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces a generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:repl-diff-dynam} provides formal detail.

@@

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side must possibly yet be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they occur. For instance, the expected utility for the sender
of choosing message $\messg$, given state $\state$ and a receiver strategy $\Rstrat$ is the
sum, for each interpretation $\state'$, of the probability that some random receiver will
choose $\state'$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and
$\state'$.

@@

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{imitation of success}
\citep[see][]{Sandholm2010:Population-Game}. The intuitive idea is the following (see
Appendix~XYZ for a derivation of the standard replicator dynamic from this update scheme).
Every now and then a random agent gets a chance to alter his behavior for one of his choice
points. Say, this is a sender who gets to ``reconsider'' his choice of message for state
$\state$. The revising agent then observes what some random agent does at $\state$, say
$\messg$, and will henceforth play $\messg$ in $\state$ with a probability given by the
expected utility of $\messg$ for state $\state$.

@@

The presence of observation and realization noise also affects imitation of successes. Here, we
focus on the main ideas. Formal detail is provided by Appendix~XYZ. Take a sender who gets to
revise behavior for state $\state$. What agents can plausibly revise by imitation is their pure
strategy, which maps perceived states onto messages. We should therefore assume that when an
agent gets to revise his strategy for the perceived state $\state$, this need not necessarily
be the actual state. Moreover, when that agent observes $\state$, another possibly
to-be-imitated agent may yet perceive a completely different state altogether. Given what that
latter agent perceives, his actual (pure) strategy will determine what he plays. In sum, to
describe how likely the potential imitator observes a message choice $\messg$, we are
interested in the conditional probability $P_o(\messg \probbar \state)$ that some other random
agent realizes $\messg$ when the first agent perceived $\state$. This
$P_o(\messg \probbar \state)$ is derived from the prior probability of states, the current
sender population behavior $\Sstrat$ and the given observation noise (see
Appendix~XYZ). Eventually, the imitating agent adopts $\messg$ as his choice for $\state$ with
probability given by the expected utility of sending $\messg$ when perceiving state
$\state$. Expected utility should, of course, take the probabilistic confusability of states
into account as well.

@@

Appendix~XYZ shows how imprecise imitation of this sort leads to mean changes in population
frequencies of choices that can be covered by the following discrete-time formulation:
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are, of course, the aforementioned
differences. Firstly, expected utility here takes stochastic confusability of states into
account. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. If there is no
observation or realization noise, $P_o(\messg \probbar \state)$ reduces to
$\sigma(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. The noisy replicator dynamic in (\ref{eq:nRD})
conservatively extends the classic case in (\ref{eq:RDDiscrete}).

@@

\subsection{Evolutionary benefits of imprecision}
The inclusion of confusability of similar states has noteworthy effects on the evolving meaning of signals.
It transpired from our results that imprecision can have further accelerating and, surprisingly, unifying effects on meaning evolution.
The unifying property of perceptual imprecision could be considered an evolutionary beneficial
side-effect. Higher imprecision can lead to higher \emph{within group expected utility},
defined as the average expected utility that each evolved language scored when playing against
an arbitrary other language obtained for the same parameter
values. Figure~\ref{fig:GroupMeasuresB} gives a representative example. The observation repeats
for other parameter values: while imprecision might decrease the communicative efficiency of
individual languages, it increases the conceptual coherence and communicative success between
independently evolving strategies. It is as if mere confusability of states imposes a
regularity constraint on evolving categories.

>>>>

The next section introduces the background against which the work presented here can be
appreciated. Section~\ref{sec:repl-diff-dynam} introduces a generalization of the replicator
dynamic which is derived from the idea that agents imitate other agents' behavior while
possibly confusing similar states. This imprecise imitation dynamic is explored in
Section~\ref{sec:exploring-rdd}. Section~\ref{sec:discussion} reflects and compares our
approach to others. Appendix~\ref{sec:formal-stuff} provides formal detail.

@@

The \emph{expected utility} of choices at each choice point is:\footnote{The notation
  $\propto$, for ``proportional to'', that is used here and hereafter means that the right-hand
  side might still need to be normalized so as to have probabilities sum to one. In general,
  writing $P(x) \propto f(x)$ for any function $f \mycolon X \rightarrow \mathds{R}$ is
  shorthand for $P(x) = \frac{f(x)}{\sum_{x' \in X} f(x')}$. }
\begin{align*}
  \EU(\messg, \state, \Rstrat) & = \sum_{\state \in \States}
  \Rstrat(\state' \probbar \messg) \myts U(\state, \state') \\
  \EU(\state', \messg, \Sstrat) & = \sum_{\state \in
    \States} P(\state \probbar \messg)  \myts
  U(\state, \state')\,, \text{\ \  where } &
P(\state \probbar \messg) & \propto \Pr(\state) \myts \Sstrat(\messg \probbar \state) \,.
\end{align*}
Expected utilities are sums, over all possible concrete outcomes, of the utilities of these
outcomes, weighted by how likely they occur. For instance, the expected utility for the sender
of choosing message $\messg$, given state $\state$ and a receiver strategy $\Rstrat$ is the
sum, for each interpretation $\state'$, of the probability that some random receiver will
choose $\state'$ given $\messg$ times the actual utility of triple $\state$, $\messg$ and
$\state'$.

@@

The replicator dynamic is an abstract population-level dynamic that describes the mean expected
change of behavioral dispositions in a population of signalers. There are several ways of
deriving the replicator dynamic from agent-level processes of behavioral adaptation. We focus
here on one of the simplest: \emph{imitation of success}
\citep[see][]{Sandholm2010:Population-Game}. The intuitive idea is the following (see
Appendix~\ref{sec:formal-stuff} for a derivation of the standard replicator dynamic from this update scheme).
Every now and then a random agent gets a chance to alter his behavior for one of his choice
points. Say, this is a sender who gets to ``reconsider'' his choice of message for state
$\state$. The revising agent then observes what some random agent does at $\state$, say
$\messg$, and will henceforth play $\messg$ in $\state$ with a probability given by the
expected utility of $\messg$ for state $\state$.

@@

The presence of observation and realization noise also affects imitation of successes. Here, we
focus on the main ideas. Formal detail is provided by Appendix~\ref{sec:formal-stuff}. Take a sender who gets to
revise behavior for state $\state$. What agents can plausibly revise by imitation is their pure
strategy, which maps perceived states onto messages. We should therefore assume that when an
agent gets to revise his strategy for the perceived state $\state$, this need not necessarily
be the actual state. Moreover, when that agent observes $\state$, another possibly
to-be-imitated agent may yet perceive a completely different state altogether. Given what that
latter agent perceives, his actual (pure) strategy will determine what he plays. In sum, to
describe how likely the potential imitator observes a message choice $\messg$, we are
interested in the conditional probability $P_o(\messg \probbar \state)$ that some other random
agent realizes $\messg$ when the first agent perceived $\state$. This
$P_o(\messg \probbar \state)$ is derived from the prior probability of states, the current
sender population behavior $\Sstrat$ and the given observation noise (see
Appendix~\ref{sec:formal-stuff}). Eventually, the imitating agent adopts $\messg$ as his choice for $\state$ with
probability given by the expected utility of sending $\messg$ when perceiving state
$\state$. Expected utility should, of course, take the probabilistic confusability of states
into account as well.

@@

Appendix~\ref{sec:formal-stuff} shows how imprecise imitation of this sort leads to mean changes in population
frequencies of choices that can be covered by the following discrete-time formulation:
\begin{align}
  \label{eq:nRD}
  \Sstrat'(\messg \probbar \state) & \propto P_o(\messg \probbar \state) \myts \EU(\messg,
  \state, \Rstrat)\,, & \Rstrat'(\state \probbar \messg) & \propto P_o(\state \probbar \messg)
  \myts \EU(\state, \messg, \Sstrat)\,.
\end{align}
\noindent This looks very much like the discrete-time formulation of the standard replicator
dynamic in (\ref{eq:RDDiscrete}), but there are, of course, the aforementioned
differences. Firstly, expected utility here takes stochastic confusability of states into
account. Secondly, where the standard replicator dynamic had probabilities
$\Sstrat(\messg \probbar \state)$ and $\Rstrat(\state \probbar \messg)$, we now have
$P_o(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ respectively. If there is no
observation or realization noise, $P_o(\messg \probbar \state)$ reduces to
$\sigma(\messg \probbar \state)$ and $P_o(\state \probbar \messg)$ reduces to
$\Rstrat(\state \probbar \messg)$. The noisy replicator dynamic in (\ref{eq:nRD})
conservatively extends the classic case in (\ref{eq:RDDiscrete}).

@@

\subsection{Evolutionary benefits of imprecision}
The inclusion of confusability of similar states has noteworthy effects on the evolving meaning of signals.
It transpired from our results that imprecision can have further accelerating and, surprisingly, unifying effects on meaning evolution.
The unifying property of perceptual imprecision could be considered an evolutionary beneficial
side-effect. A certain degree of imprecision can lead to higher \emph{within group expected utility},
defined as the average expected utility that each evolved language scored when playing against
an arbitrary other language obtained for the same parameter
values. Figure~\ref{fig:GroupMeasuresB} gives a representative example. The observation repeats
for other parameter values: while imprecision might decrease the communicative efficiency of
individual languages, it increases the conceptual coherence and communicative success between
independently evolving strategies. It is as if mere confusability of states imposes a
regularity constraint on evolving categories.
